---
title: "A hierarchical model to estimate population counts from aggregated mobile phone data"
author: "Dept. Methodology and Development of Statistical Production. Statistics Spain (INE)"
date: "29 Jan, 2018"
header_includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{amsthm}
- \usepackage[utf8]{inputnc}
output: 
  pdf_document:
    toc: true
    number_sections: true
    df_print: kable
references:
    - id: ManNacAlb15a
      title: Introduction to ecological sampling
      editor:
      - family: Manly
        given: B.F.J.
      - family: Navarro Alberto
        given: J.A.
      publisher: CRC Press
      issued:
        year: 2015
    - id: RoyDor08a
      title: Hierarchical modeling and inference in ecology
      author:
      - family: Royle
        given: J.A.
      - family: Dorazio
        given: R.M.
      publisher: Academic Press
      issued:
        year: 2008
    - id: BDA3
      title: Bayesian Data Analysis (3rd ed)
      author:
      - family: Gelman
        given: A.
      - family: Carlin
        given: J.B.
      - family: Stern
        given: H.S.
      - family: Dunson
        given: D.B.
      - family: Vehtari
        given: A.
      - family: Rubin
        given: D.B.
      publisher: CRC Press
      issued:
        year: 2013
    - id: FlaSed95a
      title: Mellin transforms and asymptotics&#58; Finite differences and Rice's integrals
      author:
      - family: Flajolet
        given: P.
      - family: Sedgewick
        given: R.
      container-title: Theoretical Computer Science 144, 101-124
      volume: 144
      page: 101-124
      issued: 
        year: 1995
    - id: GraKnuPat96a
      title: Concrete Mathematics (2nd ed.)
      author:
      - family: Graham
        given: R.L.
      - family: Knuth
        given: D.E.
      - family: Patashnik
        given: O.
      publisher: Addison-Wesley
      issued:
        year: 1996
    - id: Joh02a
      title: The curious history of Faà di Bruno's formula
      author:
      - family: Johnson
        given: W.P.
      container-title: American Mathematical Monthly 109, 217-234
      volume: 109
      page: 217-234
      issued:
        year: 2002
        month: March
    - id: BroChu03a
      title: Complex variables and applications (7th ed.)
      author:
      - family: Brown
        given: J.W.
      - family: Churchill
        given: R.V.
      issued:
        year: 2003
    - id: GraRyz07a
      title: Tables of Integrals, Series, and Products (7th ed.)
      author:
      - family: Gradshteyn
        given: I.S.
      - family: Ryzhik
        given: I.M.
      issued:
        year: 2007
    - id: Dev86a
      title: Non-uniform random variable generation
      author:
      - family: Devroye
        given: L.
      publisher: Springer
      issued:
        year: 1986
    - id: Q2016
      title: Assessing the Quality of Mobile Phone Data as a Source of Statistics
      author:
      - family: De Meersman
        given: F.
      - family: Seynaeve
        given: G.
      - family: Debusschere
        given: M.
      - family: Lusyne
        given: P.
      - family: Dewitte
        given: P.
      - family: Baeyens
        given: Y.
      - family: Wirthmann
        given: A.
      - family: Demunter
        given: C.
      - family: Reis
        given: F.
      - family: Reuter
        given: H.I. 
      container-title: Q2016 Conference
      volume: 
      page: 
      issued: 
        month: June
        year: 2016
    - id: WP5Del11
      title: Current status of access to mobile phone data in the ESS 
      author: 
        family: ESSnet on Big Data WP5
      issued:
        year: 2017
    - id: WP5IntDoc4
      title: A simple hierarchical model to estimate population counts from aggregated mobile phone data
      author: 
        family: ESSnet on Big Data WP5
      issued:
        year: 2017
       
csl: chicago-author-date.csl
---

\newtheorem{lemma}{Lemma}
\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# General introduction
This document contains a hierarchical model to estimate population counts using a combination of aggregated mobile phone data and official data both at a given time instant and along a sequence of time periods. This work completes the preceding proposal which sets the model for the initial time instant estimates (cf.\ @WP5IntDoc4). Thus, in the subsequent we show how to extend the estimates along a sequence of time instants, not just the initial one.

Again we focus only on the methodological aspects of the proposal. Software programming aspects will be dealt with in another document. We provide a full view of the proposal to offer a complete view of this methodological approach. We directly concentrate on the model leaving the general framework and motivation for the corresponding deliverables.

Now the approach can be graphically represented by figure \ref{Tool}:

\begin{figure}[htbp]
\centering
\includegraphics{Tool2.png}
\caption{\label{Tool} Schematic diagram for the output intended using mobile phone and official population data.}
\end{figure}

From the preceding stage of the whole production process with mobile phone data we get as input data for the final inference stage the number of individuals $N_{ij}^{\textrm{MNO}}(t_{0}, t_{n})$ moving from territorial cell $i$ to territorial cell $j$ in the time interval $(t_{0}, t_{n})$ according to the mobile telecommunication network. These data will be combined with official data coming from, say, the population register or another source. At the end we intend to provide as outputs (i) the probability distribution of actual individuals in each territorial cell $i$ at the initial time $t_{0}$ and (ii) the probability distribution of actual individuals at the time instants $t_{n}$ for $n=1,2,\dots$.

The two basic working assumptions are as follows:

\begin{enumerate}
\item To combine both aggregated mobile phone and official data we assume that a given time instant $t_{0}$ both population figures in each territorial cell can be equated to some extent. For the time being for ease of simplicity, we will take $N_{i}^{\textrm{Reg}}$ as a fixed quantity without a prior distribution representing uncertainty in its knowledge. Therefore $N_{i}^{\textrm{Reg}}$ will be fixed external parameters in the model. 
\item The movements of individuals across the territory (from cell to cell) is assumed to be independent of being subscribers of a given Mobile Network Operator or another.
\end{enumerate}

We propose a hierchical model supporting these hypotheses. The approach is not to think of the model as a definitive closed proposal, but as an initial step for a general methodological framework for the inference stage in which we can progressively introduce more complexities as the analysis on real data allows us to extract conclusions (e.g.\ introducing geospatial correlations among the territorial cells, dealing with the selection bias in relation with some demographic profiles -- elderly people, children, \dots).


# The complete model

Let $p_{ij}(t_{0}, t_{n})$ denote the probability for an individual to move from cell $i$ to cell $j$ in the time interval $(t_{0}, t_{n})$. Let $N_{ij}^{\textrm{MNO}}(t_{0}, t_{n})$ the number of individuals moving from cell $i$ to cell $j$ according to the network. As usual, we denote $N_{i\cdot}^{\textrm{MNO}}(t_{0})=\sum_{j=1}^{I}N_{ij}^{\textrm{MNO}}(t_{0}, t_{n})$. The complete model which we propose is specified by:

\begin{subequations}
\begin{align}
\label{eq:Model2Ini}N_{i}(t_{n}) & = \left[ N_{i}(t_{0}) + \sum_{\substack{j=1\\j\neq i}}^{I}p_{ji}(t_{0}, t_{n})N_{j}(t_{0})-\sum_{\substack{j=1\\j\neq i}}^{I}p_{ij}(t_{0}, t_{n}) N_{i}(t_{0})\right ],\quad i=1,\dots,I\\ 
\label{eq:Model2Mid}\mathbf{p}_{i\cdot}(t_{0}, t_{n})& \simeq\textrm{Dir}\left(\alpha_{i1}(t_{0}, t_{n}), \dots, \alpha_{iI}(t_{0}, t_{n})\right),\quad \mathbf{p}_{i\cdot}(t_{0}, t_{n})\perp\mathbf{p}_{j\cdot}(t_{0}, t_{n}),\quad i\neq j=1,\dots,I\\
\label{eq:Model2Fin}\alpha_{ij}(t_{0}, t_{n})&\simeq f_{\alpha ij}\left(\alpha_{ij}; \frac{N_{ij}^{\textrm{MNO}}(t_{0}, t_{n})}{N_{i\cdot}^{\textrm{MNO}}(t_{0})}\right), \quad i=1,\dots,I\\
\label{eq:Model1Ini}N_{i}^{\textrm{MNO}}(t_{0})&\simeq\textrm{Bin}\left(N_{i}(t_{0}), p_{i}(t_{0})\right),\qquad N_{i}^{\textrm{MNO}}(t_{0})\perp N_{j}^{\textrm{MNO}}(t_{0}),\quad i\neq j=1,\dots,I\\
N_{i}(t_{0})&\simeq\textrm{Po}\left(\lambda_{i}(t_{0})\right),\qquad N_{i}(t_{0})\perp N_{j}(t_{0}),\quad i\neq j=1,\dots,I\\
p_{i}(t_{0})&\simeq\textrm{Beta}\left(\alpha_{i}(t_{0}),\beta_{i}(t_{0})\right),\qquad p_{i}(t_{0})\perp p_{j}(t_{0})\quad i\neq j=1,\dots,I\\
\left(\alpha_{i}(t_{0}), \beta_{i}(t_{0})\right)&\simeq \frac{f_{ui}\left(\frac{\alpha_{i}}{\alpha_{i}+\beta_{i}}; \frac{N_{i}^{\textrm{MNO}}(t_{0})}{N_{i}^{\textrm{REG}}(t_{0})}\right)\cdot f_{vi}\left(\alpha_{i}+\beta_{i}; N_{i}^{\textrm{REG}}(t_{0})\right)}{\alpha_{i}+\beta_{i}},\quad (\alpha_{i}(t_{0}),\beta_{i}(t_{0}))\perp(\alpha_{j}(t_{0}),\beta_{j}(t_{0})),\quad i\neq j=1,\dots,I\\
\label{eq:Model1Fin}\lambda_{i}(t_{0})&\simeq f_{\lambda i}(\lambda_{i}; N_{i}^{\textrm{REG}}(t_{0}))\quad (\lambda_{i}(t_{0}) > 0,\quad \lambda_{i}(t_{0})\perp\lambda_{j}(t_{0})), \quad i=1,\dots,I,
\end{align}
\end{subequations}

\noindent where 

\begin{itemize}
\item $[\cdot]$ denotes the nearest integer function;
\item $f_{\alpha ij}$ stands for the probability density function of the parameters $\alpha_{ij}$. The notation $f_{\alpha ij}\left(\alpha_{ij}; \frac{N_{ij}^{\textrm{MNO}}(t_{0}, t_{n})}{N_{i\cdot}^{\textrm{MNO}}(t_{0})}\right)$ is meant to indicate that $\frac{N_{ij}^{\textrm{MNO}}(t_{0}, t_{n})}{N_{i\cdot}^{\textrm{MNO}}(t_{0})}$ should be taken as the mode of the density function;
\item $f_{ui}$ stands for the probability density function of the parameter $u$ (see the preceding document on the model) in cell $i$ with mode $\frac{N_{i}^{\textrm{MNO}}(t_{0})}{N_{i}^{\textrm{REG}}(t_{0})}$;
\item $f_{vi}$ stands for the probability density function of the parameter $v$ (see the preceding document on the model) in cell $i$ with mode $N_{i}^{\textrm{REG}}(t_{0})$;
\item $f_{\lambda i}$ stands for the probability density function of the parameter $\lambda$ (see the preceding document on the model) in cell $i$ with mode $N_{i}^{\textrm{REG}}(t_{0})$. 
\end{itemize}

Equations \eqref{eq:Model1Ini} to \eqref{eq:Model1Fin} are taken from the model in the preceding document just adding explicitly the time dependence. Therefore, their interpretation remains the same. Equations \eqref{eq:Model2Ini}, \eqref{eq:Model2Mid}, \eqref{eq:Model2Fin} take care of the time evolution of the estimates. Their interpretation is straightforward.

Equation \eqref{eq:Model2Ini} states that the number of individuals in a cell $i$ at time $t_{n}$ equals the initial number of individuals plus those arriving from other cells in the given time interval minus those leaving for other cells in the same time interval. The number of individuals arriving and leaving are estimated using the transition probability among cells.

Next, to estimate these transition probabilities we model them for a given cell $i$ as a multivariate random variable with a Dirichlet distribution with parameters $\alpha_{i1},\dots,\alpha_{iI}$. These parameters, in turn, are given unimodal prior distributions $f_{\alpha_{ij}}$ with mode in $\frac{N_{ij}^{\textrm{MNO}}}{N_{i\cdot}^{\textrm{MNO}}}$ according to our second working assumption.

# Computations

Ultimately we need to compute the probability functions $\mathbb{P}\left(N_{i}(t_{n})\big|N^{\textrm{MNO}}(t_{0}, t_{1})\right)$ for each cell $i$, which will allow us to choose an estimator as, e.g., the posterior mean, posterior median or posterior mode.

The computation is conducted empirically in three steps:

\begin{enumerate}
\item The initial population value $N_{i}(t_{0})$ is generated for all cells $i=1,\dots, I$ according to the model using $N_{i\cdot}^{\textrm{MNO}}(t_{0})$ as input data and choosing weakly informative priors $f_{ui}$, $f_{vi}$ and $f_{\lambda i}$.
\item A transition probability matrix $[p_{ij}(t_{0}, t_{n})]$ is generated according to the model using $N^{\textrm{MNO}}(t_{0}, t_{n})$ as input data and choosing weakly informative priors $f_{\alpha ij}$.
\item These generated quantities are used in formula \eqref{eq:Model2Ini} to generate $N_{i}(t_{1})$ for all cells $i=1,\dots,I$.
\end{enumerate}

Following these steps we can generate an empirical posterior distribution of values $N_{i}(t_{n})$ for each cell $i$. Then we can use these distributions to provide a point estimate according to its mean, median, or mode. 

# A toy example

Again let us simulate a toy population to illustrate this approach. We generate a population with a territory divided into $12$ cells. At $4$ succesive time intervals the individuals follow their own trajectories so that the population in each cell evolves according to actual transition matrices $N_{ij}(t_{n-1}, t_{n})$, $n=1,2,3,4$. At the initial time instant $t_{0}$ we also have the population register figures for each cell $N_{i}^{\textrm{Reg}}(t_{0})$ which can be equated to the initial real population $N_{i}(t_{0})$, although they are not completely equal due to nonsampling errors. Data used in this illustration can be found in URL.

We also generate the transition matrices for each time interval $n=1,2,3,4$ for the individuals $N_{ij}^{\textrm{MNO}}(t_{n-1}, t_{n})$ detected with the network.

We proceed as sketched above. Firstly, we generate the estimates $N_{i}(t_{0})$ for the initial time instant, as conducted in the preceding document.

```{r estimN0, eval=TRUE, cache=TRUE, echo = FALSE}
library(data.table)
library(ggplot2)
Sys.setenv(PATH = paste(Sys.getenv("PATH"), "C:/Rtools/bin/",
            "C:/Rtools/mingw_64/bin", sep = ";"))
Sys.setenv(BINPREF = "C:/Rtools/mingw_64/bin/")
source('N:/UDMTD/UDTMDCOM/BigData/MetodologiaDatosTelefonia/Functions23.R')
Sim.PopMob <- readRDS('N:/UDMTD/UDTMDCOM/BigData/MetodologiaDatosTelefonia/Sim.PopMob.rds')
InitialPop <- Sim.PopMob[ID_T == 1 & ID_CELL_FIN == ID_CELL_INI]
Scale <- 1e3
nMNO_ini <- InitialPop[['N_MNO_1']] / Scale
nReg <- InitialPop[['N_REG']] / Scale
u0 <- InitialPop$N_MNO_1 / InitialPop$N_REG
fu <- lapply(u0, function(u){
  
  umin <- max(0, u - 0.10 * u)
  umax <- min(1, u + 0.10 * u)
  output <- list('triang', xMin = umin, xMax = umax, xMode = u)
  return(output)
  
})
v0 <- nReg
fv <- lapply(v0, function(u){
  
  umin <- max(0, u - 0.10 * u)
  umax <- u + 0.10 * u
  output <- list('triang', xMin = umin, xMax = umax, xMode = u)
  return(output)
  
})
alpha <- 1 / 0.1**2 - 1
flambda <- lapply(v0, function(v){list('gamma', shape = 1 + alpha, scale = v / alpha)})

nSim <- 10000
N0cells <- rN0(nSim, nMNO_ini, nReg, fu, fv, flambda, scale = Scale)
ggplot(N0cells, aes(x = N0)) + 
  geom_histogram(binwidth = 50) +
  geom_vline(aes(xintercept = nReg), color = 'red') +
  facet_grid(factor(cellID) ~ ., scales="free") +
  xlab('Posterior populations per cell') + ylab('Counts (different scales)\n') +
  ggtitle(paste0('Simulated posterior populations for each cell\n')) +
  theme(plot.title = element_text(face = 'bold', size = 14, hjust = 0.5), axis.text.y = element_blank())


```
Now we can provide two kind of time evolutions for the population in each cell. On the one hand, we can generate simulated populations conditioned upon their estimated initial size $\widehat{N}_{i}(t_{0})$. On the other hand, we can provide these simulated populations unconditioned upon their estimated initial size but starting from the input data themselves (thus uncertainty in the initial population estimate is included).

In the first case, taking the mean of the preceding populations as estimates for the initial population of each cell we obtain the following evolving simulated populations in each cell (in dashed lines the assumed true values of the simulated population):

```{r estimNtcondN0, dependson = 'estimN0', eval=TRUE, cache=TRUE, echo = FALSE}
N0 <- N0cells[, list(Mean = round(mean(N0))), by = 'cellID'][['Mean']]
nT <- length(unique(Sim.PopMob$ID_T)) - 1
matList <- lapply(2:(nT + 1), function(i){
  
  output <- dcast(Sim.PopMob[ID_T == i], ID_CELL_INI ~ ID_CELL_FIN, value.var = 'N_MNO_1')
  output[, ID_CELL_INI := NULL]
  return(output)
  
})
DistributNames <- rep('triang', 12)
Variation <- rep(list(list(cv = 0.10)), 12)
NtcondN0 <- rbindlist(lapply(seq(along = matList), function(i){
  
  output <- rNtcondN0(nSim, N0, as.matrix(matList[[i]]),  DistributNames, Variation)
  output <- as.data.table(output)
  setnames(output, as.character(1:dim(output)[2]))
  output[, sim := 1:.N]
  output <- melt(output, id.vars = 'sim',  variable.name = 'Cell', value.name = 'N0')
  output[, Time := factor(i)]
  return(output)
}))

N0True <- readRDS('N:/UDMTD/UDTMDCOM/BigData/MetodologiaDatosTelefonia/TruePop.rds')
N0True <- N0True[Time != 1][, Time := factor(Time - 1)]
ggplot(NtcondN0, aes(x = N0, fill = Time)) + 
  geom_histogram(binwidth = 50, alpha=.5, position="identity") + 
  geom_vline(data = N0True, aes(xintercept = N0True, color = Time), linetype = 'dashed', size = 0.5) + 
  facet_grid(factor(Cell) ~ ., scales="free") +
  xlab('\n Posterior populations per cell') + ylab('Counts (different scales)\n') +
  ggtitle(paste0('Simulated posterior populations for each cell\n (Conditioned on initial population)')) +
  theme(plot.title = element_text(face = 'bold', size = 14, hjust = 0.5), axis.text.y = element_blank())  
  

```
By comparing both figures we can detect how the uncertainty in the estimation of the initial population size propagates along the evolving estimates of the population size of each cell. In particular, in cells 6 and 7 we can observe how the uncertainty in the simulated initial populations provide rather inaccurate estimates for later time periods. 

Using again the posterior mean as estimator of the population in each cell, we have the following evolutions:

```{r estimEvolNtcondN0, dependson = 'estimNtcondN0', eval=TRUE, cache=TRUE, echo = FALSE}
N0dt <- data.table(N = N0, Time = 0, Cell = seq(along=N0))
seriesNt <- NtcondN0[, list(N = round(mean(N0))), by = c('Time', 'Cell')]
setcolorder(N0dt, names(seriesNt))
seriesNt <- rbindlist(list(N0dt, seriesNt))
NTrue <- readRDS('N:/UDMTD/UDTMDCOM/BigData/MetodologiaDatosTelefonia/TruePop.rds')
NTrue[, Time := factor(Time - 1)]
setnames(NTrue, 'N0True', 'trueN')
seriesNt <- merge(seriesNt, NTrue)
seriesNt <- melt(seriesNt, id.vars = c('Time', 'Cell'), value.name = 'Population', variable.name = 'Series')
ggplot(seriesNt, aes(x = Time, y = Population, color = Series, group = Series)) + 
  geom_line() + 
  facet_grid(factor(Cell) ~ ., scales="free") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
  xlab('\n Time Period') + ylab('Population (different scales; watch out the origins)\n') +
  ggtitle(paste0('Estimated populations for each cell\n (Conditioned on initial population)')) +
  scale_color_manual(labels = c(expression(widehat(N)), expression(N^{(0)})), values = c('red', 'blue')) +
  theme(plot.title = element_text(face = 'bold', size = 14, hjust = 0.5), axis.text.y = element_text(size = 5))
```

In terms of the relative bias, this comparison can be depicted as follows:


```{r realBiasEvolNtcondN0, dependson = 'estimEvolNtcondN0', eval=TRUE, cache=TRUE, echo = FALSE}
seriesNt.dcast <- dcast(seriesNt, Time + Cell ~ Series, value.var = 'Population')
seriesNt.dcast[, relBias := round((N - trueN) / trueN * 100, 2)][, (c('N', 'trueN')) := NULL]
seriesNt.melt <-  melt(seriesNt.dcast, id = c('Time', 'Cell'))
ggplot(seriesNt.melt, aes(x = Time, y = value, group = variable, color = variable)) + 
  geom_line() + 
  facet_grid(Cell ~ .) +
  geom_hline(yintercept = 0, size = 0.3, linetype = 'dotted') + 
  xlab('\n Time Period') + ylab('Relative bias\n') +
  ggtitle(paste0('Relative bias of estimated population size\n (conditioned on initial population)')) +
  theme(plot.title = element_text(face = 'bold', size = 12, hjust = 0.5), axis.text.y = element_text(size = 5), legend.position="none")

```

Complementarily, we can produce estimates unconditioned on the initial estimate of the population in each cell. Now the uncertainty in the estimation of the initial population of each cell is incorporated into the estimation process for later time periods.

```{r estimNt, dependson = 'estimNtcondN0', eval=TRUE, cache=TRUE, echo = FALSE}

nT <- length(unique(Sim.PopMob$ID_T)) - 1
matList <- lapply(2:(nT + 1), function(i){
  
  output <- dcast(Sim.PopMob[ID_T == i], ID_CELL_INI ~ ID_CELL_FIN, value.var = 'N_MNO_1')
  output[, ID_CELL_INI := NULL]
  output <- output / Scale
  return(output)
  
})
DistributNames <- rep('triang', 12)
Variation <- rep(list(list(cv = 0.10)), 12)
Nt <- rbindlist(lapply(seq(along = matList), function(i){
  
  output <- rNt(nSim, as.matrix(matList[[i]]),  nReg, fu, fv, flambda, DistributNames, Variation, scale = Scale)
  output[, Time := factor(i)]
  return(output)
}))
NTrue <- readRDS('N:/UDMTD/UDTMDCOM/BigData/MetodologiaDatosTelefonia/TruePop.rds')
NTrue[, Time := factor(Time - 1)]
setnames(Nt, c('sim', 'Cell', 'N', 'Time'))
seriesNt <- Nt[, list(N = round(mean(N))), by = c('Time', 'Cell')]
Nt0dt <- data.table(N = N0, Time = 0, Cell = seq(along = N0))
setcolorder(Nt0dt, c('Time', 'Cell', 'N'))
seriesNt <- rbindlist(list(Nt0dt, seriesNt))
seriesNt[, Cell := as.factor(Cell)]
seriesNt <- merge(seriesNt, NTrue)
setnames(seriesNt, 'N0True', 'NTrue')
seriesNt <- melt(seriesNt, id.vars = c('Time', 'Cell'), value.name = 'Population', variable.name = 'Series')
ggplot(seriesNt, aes(x = Time, y = Population, color = Series, group = Series)) + 
  geom_line() + 
  facet_grid(factor(Cell) ~ ., scales="free") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
  xlab('\n Time Period') + ylab('Population (different scales; watch out the origins)\n') +
  ggtitle(paste0('Estimated populations for each cell\n (Unconditioned on initial population)')) +
  scale_color_manual(labels = c(expression(widehat(N)), expression(N^{(0)})), values = c('red', 'blue')) +
  theme(plot.title = element_text(face = 'bold', size = 14, hjust = 0.5), axis.text.y = element_text(size = 5))
  
  
```
In terms of the relative bias, this comparison can be depicted as follows:


```{r realBiasEvolNt, dependson = 'estimNt', eval=TRUE, cache=TRUE, echo = FALSE}
seriesNt.dcast <- dcast(seriesNt, Time + Cell ~ Series, value.var = 'Population')
seriesNt.dcast[, relBias := round((N - NTrue) / NTrue * 100, 2)][, (c('N', 'NTrue')) := NULL]
seriesNt.melt <-  melt(seriesNt.dcast, id = c('Time', 'Cell'))
ggplot(seriesNt.melt, aes(x = Time, y = value, group = variable, color = variable)) + 
  geom_line() + 
  facet_grid(Cell ~ .) +
  geom_hline(yintercept = 0, size = 0.3, linetype = 'dotted') + 
  xlab('\n Time Period') + ylab('Relative bias\n') +
  ggtitle(paste0('Relative bias of estimated population size\n (conditioned on initial population)')) +
  theme(plot.title = element_text(face = 'bold', size = 12, hjust = 0.5), axis.text.y = element_text(size = 5), legend.position="none")

```

# First conclusions

Apparently, the model seems to work on these toy data. However there are many details to be tackled on before claiming such a conclusion:

\begin{enumerate}
\item The model must be checked against larger and more complex data (more cells and more time intervals). The process of data generation must be explicitly spelled out to assess about how realistic they are. This will be undertaken in the deliverable.
\item It is expected that the higher the complexity of the data is, the more computationally demanding the model will be. We will have to check the computational capabilities of the programmed functions.
\item No mention to accuracy (apart from the relative bias) and its estimation has been included. This will be an essential part of the quality assessment and will be undertaken in the last deliverable.
\item As the data complexity increases, we will certainly need a visualization tool to include population estimates of all cells for all time periods (possibly maps, grids, heat maps, etc.).
\item The model ultimately must be tested against real data compiled in the SGA-1 by WP members. In this case we do not have the ground truth figures to compare with and we must rely on the preceding performance and expert knowledge.
\item Notice that the model itself (not the data) can gain in complexity by adding new methodological elements:
    \begin{itemize}
      \item The register population $N_{i}^{\textrm{Reg}}$ is not modelled at all. We can also assume a prior distribution for these variables reflecting the uncertainty (due to nonsampling errores) in the register figures.
      \item No spatial correlation is introduced in the model between the cells. These correlations can introduce more realistically this feature.
      \item No time correlation is introduced in the model between time periods. Time series techniques (or space-time modelling techniques) can in principle improve the model.
      \item No bias selection analysis has been conducted. Techniques as Heckman bias correction technique can be also another element to take into consideration.
    \end{itemize}
\end{enumerate}

Thus, this proposal is indeed a first simple model which can be further potentially adapted to more realistic situations as the analysis with real data will suggest.


# References
